Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-10-12 experiments.tex
AA: implementati i suggerimenti sull'espressione inglese del reviewer II

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. At the best of our knowledge, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Pareto Sets.  Note that, for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis.  We also exclude other approaches (e.g.  dependency
based~\cite{givargis_tvlsi02}) that make use of some a-priori
knowledge about the role and meaning of each parameter, since our aim
is evaluate the capacity of focusing on interesting parameter regions
without any external judgement or help based on heuristics.  Finally,
it should be pointed out that in this introductory work we will
explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of
MOGA, remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03}
for a detailed analysis on how it differs from other genetic approaches
and what could be expected from similar multi-objective strategies.

@@

Set of Figures~\ref{fig:pareto_fronts_100}
and~\ref{fig:pareto_fronts_500} show the resulting Pareto fronts for two
different scenarios of 100 and 500 simulations budget. When possible,
similar algorithm parameters have been adopted. In particular a
population size of 30 has been chosen for the genetic approach (GA), corresponding to the
same value for the $K$ of PS. The number of generations (or era)
has been chosen so that the only the given amount of simulations is
actually performed. Thus, when accounting these simulations only unique simulations
has been considered, removing repetitions and unfeasible
configurations.

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 100 configurations.\label{fig:pareto_fronts_100}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_100.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_100.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int100.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_100.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_100.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_100.eps} 
  \end{tabular}}

\end{table}

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly dependent upon the application being considered.
These results are obtained in about 5 eras and demonstrate that with a so limited budget PS can compete with
genetic approach while even discovering very different Pareto fronts. 
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective.
A comparison between the variation ranges of both power dissipation
and execution time, obtained by PS and GA for different benchmarks is
shown in Fig.~\ref{fig:dispersion}(a) and (b) respectively.  As it can
be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than the one provided by a GA
exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

\begin{table}
  \tbl{(a)(b) Variation range obtained by PS and GA for different benchmarks
  and average dispersion absolute dispersion (c). \comment{AA: manca il titolo nella figura (c)}
  \label{fig:dispersion}}{
  \begin{tabular}{c}
    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
    (a) \\
    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
    (b) \\
    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
    (c) 
  \end{tabular}}
\end{table}
The average normalised absolute dispersion error measures
the average absolute difference between the distribution of points in
the objective space and an ideal distribution in which the points are
uniformly distributed over the objective space. Formally, let $O$ be
the image, in the objective space, of the configurations visited by
the design space exploration. The generic element of $O$ (\ie, a
solution) is a pair $(p,t)$ where $p$ and $t$ are the average power
and execution time, respectively. The two-dimensional objective space
is then partitioned by a $M_x \times M_y$ mesh. For each tile $T_i,
\ i=1, 2, \ldots, M_xM_y$ of the mesh, let $N_i$ be the number
of points in $O$ which fall in $T_i$. The average absolute error, $E_i$, for
$T_i$ is the absolute value of the difference between $N_i$ and the
ideal number of solutions, $\overline{N}$, which should fall in $T_i$
in case of uniform distribution. Such $\overline{N}$ can be simply
computed as the ratio between the cardinality of $O$ and the number of
tiles. Thus,
\[ E_i = |N_i - \overline{N}|, \]
where $\overline{N} = |O| / (M_x M_y)$. The average
normalised absolute dispersion error ($ANADE$) is the average of $E_i$
normalised to the maximum absolute error $E_{max}$:
\[ ANADE = \frac{\sum_{i=1}^{M_xM_y} E_i/(M_xM_y)}{E_{max}}, \]
where $E_{max}$ can be computed as the average absolute error in the worst
case in which all the solutions fall in a single tile:
\[ E_{max} = \frac{(M_x M_y - 1) \overline{N} + |\overline{N} -
    |O|| }{M_x M_y}. \]

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
is worth pointing out how a definitive claim about the superiority of
the PS behaviour is strictly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevertheless, the proposed algorithm introduces a new
perspective on the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software design space exploration.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm
(MOGA)~\cite{knowles_techrep06}. To the best of our knowledge, genetic
approaches in general still represent the best compromise between
efficiency (i.e. time required for exploration) and accuracy of the
reported Pareto Sets.  Note that, for the reasons described above,
mono-objective approaches have been explicitly excluded from this
analysis.  We also exclude other approaches (e.g.  dependency
based~\cite{givargis_tvlsi02}) that make use of some a-priori
knowledge about the role and meaning of each parameter, since our aim
is evaluate the capacity of focusing on interesting parameter regions
without any external judgement or help based on heuristics.  Finally,
it should be pointed out that in this introductory work we will
explicitly use the \emph{SPEA}~\cite{zitzler_eurogen01} variant of
MOGA, remanding to works like~\cite{zitzler_ec00}~\cite{zitzler_tec03}
for a detailed analysis on how it differs from other genetic approaches
and what could be expected from similar multi-objective strategies.

@@

\figR{pareto_fronts_100} and \figR{pareto_fronts_500} show the resulting Pareto fronts for two
different scenarios of 100 and 500 simulations budget. When possible,
similar algorithm parameters have been adopted. In particular a
population size of 30 has been chosen for the genetic approach (GA), corresponding to the
same value for the $K$ of PS. The number of generations (or era)
has been chosen so that the only the given amount of simulations is
actually performed. Thus, when accounting these simulations only unique simulations
has been considered, removing repetitions and unfeasible
configurations.

@@

\begin{figure}
  \figLC{pareto_fronts_100}{Pareto fronts found by PS and GA for a fixed budget of 100 configurations. \comment{AA: nelle due figure centrali la linea di GA e' continua, mentre in tutte le altre figure e' tratteggiata. Il revisore ci chiede di uniformare. Non so come riottenere le figure. Potresti pensarci tu?}}
  \begin{center}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_100.eps}
    \includegraphics[width=0.30\textwidth]{pictures/bmm_100.eps}
    \includegraphics[width=0.30\textwidth]{pictures/fir_int100.eps}
    \includegraphics[width=0.30\textwidth]{pictures/mm_100.eps}
    \includegraphics[width=0.30\textwidth]{pictures/sha_100.eps}
    \includegraphics[width=0.30\textwidth]{pictures/wave_100.eps} 
  \end{center}
\end{figure}

\begin{figure}
  \figLC{pareto_fronts_500}{Pareto fronts found by PS and GA for a fixed budget of 500 configurations.\comment{AA: nelle due figure centrali la linea di GA e' continua, mentre in tutte le altre figure e' tratteggiata. Il revisore ci chiede di uniformare. Non so come riottenere le figure. Potresti pensarci tu?}}
  \begin{center}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps}
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps}
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps}
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps}
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps}
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{center}
\end{figure}
Low budget Pareto sets are shown in \figR{pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly dependent upon the application being considered.
These results are obtained in about 5 eras and demonstrate that with a so limited budget PS can compete with
genetic approach while even discovering very different Pareto fronts. 
Considering the extended simulation budget as depicted in
\figR{pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective.
A comparison between the variation ranges of both power dissipation
and execution time, obtained by PS and GA for different benchmarks is
shown in \figR{dispersion}(a) and (b) respectively.  As it can
be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than the one provided by a GA
exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

\begin{figure}
  \figLC{dispersion}{(a)(b) Variation range obtained by PS and GA for different benchmarks and average dispersion absolute dispersion (c). \comment{AA: manca il titolo nella figura (c)} }
  \begin{center}
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} }
  \end{center}
\end{figure}

The average normalised absolute dispersion error measures
the average absolute difference between the distribution of points in
the objective space and an ideal distribution in which the points are
uniformly distributed over the objective space. Formally, let $O$ be
the image, in the objective space, of the configurations visited by
the design space exploration. The generic element of $O$ (\ie, a
solution) is a pair $(p,t)$ where $p$ and $t$ are the average power
and execution time, respectively. The two-dimensional objective space
is then partitioned by a $M_x \times M_y$ mesh. For each tile $T_i,
\ i=1, 2, \ldots, M_xM_y$ of the mesh, let $N_i$ be the number
of points in $O$ which fall in $T_i$. The average absolute error, $E_i$, for
$T_i$ is the absolute value of the difference between $N_i$ and the
ideal number of solutions, $\overline{N}$, which should fall in $T_i$
in case of uniform distribution. Such $\overline{N}$ can be simply
computed as the ratio between the cardinality of $O$ and the number of
tiles. Thus,
\[ E_i = |N_i - \overline{N}|, \]
where $\overline{N} = |O| / (M_x M_y)$. The average
normalised absolute dispersion error ($ANADE$) is the average of $E_i$
normalised to the maximum absolute error $E_{max}$:
\[ ANADE = \frac{\sum_{i=1}^{M_xM_y} E_i/(M_xM_y)}{E_{max}}, \]
where $E_{max}$ can be computed as the average absolute error in the worst
case in which all the solutions fall in a single tile:
\[ E_{max} = \frac{(M_x M_y - 1) \overline{N} + |\overline{N} -
    |O|| }{M_x M_y}. \]

\figR{dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of \figR{dispersion}(a) and
(b) confirm the of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
is worth pointing out how a definitive claim about the superiority of
the PS behaviour is strictly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevertheless, the proposed algorithm introduces a new
perspective on the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software design space exploration.

### 2014-10-12 introduction.tex
AA: implementati i suggerimenti sull'espressione inglese del reviewer II

In these cases a Design Space
Exploration (DSE) strategy is required, i.e. an algorithm that selects
only a tractable subset of all possible configurations.
The final result of a design space exploration is a subset of the
configurations called \emph{Pareto set} (\cite{pareto}, see also
\defR{pareto-set}). The corresponding values of the objectives for
each configuration of the Pareto Set is called \emph{Pareto front}:
it represents the trade-offs between objectives, i.e.  for each point
of the Pareto front there is no other configuration performing better
for all the objectives considered (see Figure~\ref{fig:psos}). Of course this is only a first
high-level system simulation, thus these few promising
configurations will need to be further evaluated in a next step of the design
flow using a more accurate low level simulation (e.g. register
transfer level).

@@

In this work we present PS, a multiobjective design exploration
algorithm that introduces the concept of \emph{interesting} and
\emph{uninteresting regions} of the configuration space.  The main
idea can be simplified as follows: at each iteration of the algorithm,
the configuration space is divided in regions. A score is given to
each region based on the \emph{innovation} provided by its
configurations in the Pareto front. The regions with high score are
considered the most \emph{interesting} and more simulations are
distributed there rather than in the rest of the configuration space.
Several geometrical transformations (splitting, merging etc...) are
involved in the Parameter Space (aka configuration space), hence
the name PS for the algorithm.

@@

The paper is organized as follows. First, \secR{Related-work} places our work in the context of related effort.
Then we give a theoretical formulation of the algorithm. In particular, in \secR{statement_of_the_problem} we provide formal definitions of the concepts and the operations that will be used to describe PS algorithm. In \secR{algorithm} we give the theoretical description of PS. 
Finally in Section~\ref{sec:ee} we show a case
study involving the exploration of the hardware/software parameters of
a Very Long Instruction Word (VLIW) architecture, performing a
qualitative and quantitative comparison of PS against the state-of-art
of multiobjective genetic based approaches.

@@

Abraham, Rau and Schreiber proposed in \cite{santosh_hptr00} to decompose
the system under evaluation into components that interact minimally
with each other. Pareto sets for each component are found and, provided
that ``monotonicity'' exists, the complete Pareto set is computed
merging the component Pareto sets. Roughly speaking (see section 4
of the same paper for more details), monotonicity property guarantees
that the best system can be obtained as a composition of the best
components. This approach would perfectly work if all the components
were perfectly isolated, i.e., if there were no dependencies among
them. But real scenarios seldom if ever
expose monotonicity property thus resulting in inaccuracies, as stated in the same paper).

>>>>

In these cases a Design Space
Exploration (DSE) strategy is required, i.e. an algorithm that selects
only a tractable subset of all possible configurations.
The final result of a design space exploration is a subset of the
configurations called \emph{Pareto set} (\cite{pareto}, see also
\defR{pareto-set}). The corresponding values of the objectives for
each configuration of the Pareto Set is called \emph{Pareto front}:
it represents the trade-offs between objectives, i.e.  for each point
of the Pareto front there is no other configuration performing better
for all the objectives considered (see \figR{psos}). Of course this is only a first
high-level system simulation, thus these few promising
configurations will need to be further evaluated in a next step of the design
flow using a more accurate low level simulation (e.g. register
transfer level).

@@

In this work we present PS, a multiobjective design exploration
algorithm that introduces the concept of \emph{interesting} and
\emph{uninteresting regions} of the configuration space.  The main
idea can be simplified as follows: at each iteration of the algorithm,
the configuration space is divided in regions. A score is given to
each region based on the \emph{innovation} provided by its
configurations in the Pareto front. The regions with high score are
considered the most \emph{interesting} and more simulations are
distributed there rather than in the rest of the configuration space.
Several geometrical transformations (splitting, merging etc.) are
involved in the Parameter Space (aka configuration space), hence
the name PS for the algorithm.

@@

The paper is organized as follows. First, \secR{Related-work} places our work in the context of related effort.
Then we give a theoretical formulation of the algorithm. In particular, in \secR{statement_of_the_problem} we provide formal definitions of the concepts and the operations that will be used to describe PS algorithm. In \secR{algorithm} we give the theoretical description of PS. 
Finally in \secR{ee} we show a case
study involving the exploration of the hardware/software parameters of
a Very Long Instruction Word (VLIW) architecture, performing a
qualitative and quantitative comparison of PS against the state-of-art
of multiobjective genetic based approaches.

@@

Abraham, Rau and Schreiber proposed in \cite{santosh_hptr00} to decompose
the system under evaluation into components that interact minimally
with each other. Pareto sets for each component are found and, provided
that ``monotonicity'' exists, the complete Pareto set is computed
merging the component Pareto sets. Roughly speaking (see section 4
of the same paper for more details), monotonicity property guarantees
that the best system can be obtained as a composition of the best
components. This approach would perfectly work if all the components
were perfectly isolated, i.e., if there were no dependencies among
them. But real scenarios rarely expose monotonicity property thus resulting in inaccuracies, as stated in the same paper).

### 2014-10-12 paper.tex
AA: implementati i suggerimenti sull'espressione inglese del reviewer II

%\newcommand{\comment}[1]{\textcolor{red}{\bf #1}}
\newcommand\comment[1]{} % When you enable this lines, all the notes will disappear

@@

\section{Conclusions}
In this work we presented PS, a multiobjective strategy which 
introduces the concept of Parameter Space representation of Pareto
Front. A case study of a VLIW architecture involving strong
hardware/software dependencies has been analyzed to evaluate
effectiveness with different amount of simulations budget.
A qualitative/quantitative comparison against a widespread multiobjective genetic
approach showed how the proposed strategy can result in a different
distribution of Pareto points which trades solutions granularity with
improved Pareto extension.

>>>>

\newcommand{\comment}[1]{\textcolor{red}{\bf #1}}
%\newcommand\comment[1]{} % When you enable this lines, all the notes will disappear

@@

\secL{Conclusions}
In this work we presented PS, a multiobjective strategy which 
introduces the concept of Parameter Space representation of Pareto
Front. A case study of a VLIW architecture involving strong
hardware/software dependencies has been analyzed to evaluate
effectiveness with different amount of simulations budget.
A qualitative/quantitative comparison against a widespread multiobjective genetic
approach showed how the proposed strategy can result in a different
distribution of Pareto points which trades solutions granularity with
improved Pareto extension.
